Predicting Survival Probability of NASA Aircraft Engines

### 2025-04-16 index.qmd
Update 4/15 pt3

After determining the model to use with the covariates, the CPH
assumptions need to be tested to confirm the model is a valid CPH model.
The model may have a high concordance index, but it may not be valid for
CPH to use to predict the survival probability. If any of the
assumptions for the model are not true, the model will need to be
modified.

@@

The survival times for each individual engine in the training data are
independent of each other. The failure of one engine does not impact the
failure of another so this assumption is met.

@@

The training dataset does not contain censored data as the documentation
available at the dataset’s source clarifies all engines in the training
data have experienced the desired event of failure at the end of each
time series. As such, this assumption is met.

@@

The coefficients above can be used to calculate the hazard rate
[@eq-hazardrate], the cumulative hazard value[@eq-cumulativehr], and the
survival probability [@eq-survprob]. The hazard rate at time $t$ will be
calculated for each engine at the engines max time value. The cumulative
hazard value is the total hazard accumulated for the subject over time
and is used to estimate the risk for the subject at a point in time. The
engines that have higher hazard rates or higher cumulative hazard values
are at a greater risk of experiencing failure. The below table provides
the hazard rate and suvivial probability for the first 10 engines in the
dataset.

@@

The implications of an engine failure are severe, as failure can lead to
injury or death and the loss of millions of dollars. As such,
understanding when an engine will fail allows for maintenance to occur
before negative impacts occur. While CPH does not allow for the
prediction of specific failure times, it does allow for the creation
confidence intervals which can be used to gain an understanding of when
engines are most likely to fail. The table below displays the percent of
survival at the lower end of the confidence interval to exercise
caution.

>>>>

After determining the model to use with the covariates, the CPH
assumptions need to be tested to confirm the model is a valid CPH model.
The model may have a high concordance index, but it may not be valid for
CPH to use to predict the survival probability. If any of the
assumptions for the model are not met, the model will need to be
modified.

@@

The survival times for each individual engine in the data are
independent of each other. The failure of one engine does not impact the
failure of another so this assumption is met.

@@

While the data that was pulled from the training set does not include
censored data, all of the data from the testing dataset is censored as
those engines had not experienced failure in the study time period.
However, it is known that those engines will experience failure at some
point and knowing those engines are censored does not provide any new
information. As such, this assumption is met.

@@

The coefficients above can be used to calculate the hazard rate
[@eq-hazardrate], the cumulative hazard value[@eq-cumulativehr], and the
survival probability [@eq-survprob]. The hazard rate at time $t$ will be
calculated for each engine at the engines max time value. The cumulative
hazard value is the total hazard accumulated for the subject over time
and is used to estimate the risk for the subject at a point in time. The
engines that have higher hazard rates or higher cumulative hazard values
are at a greater risk of experiencing failure. The below table provides
the hazard rate and survival probability for the first 10 engines in the
dataset.

@@

The implications of an engine failure are severe, as failure can lead to
injury or death and the loss of millions of dollars. As such,
understanding when an engine will fail allows for maintenance to occur
before negative impacts occur. While CPH does not allow for the
prediction of specific failure times, it does allow for the creation of
confidence intervals which can be used to gain an understanding of when
engines are most likely to fail. The table below displays the percent of
survival at the lower end of the confidence interval to exercise
caution.

### 2025-04-16 slides.qmd
Update 4/15 pt3

The two common model visualizations are:

::: incremental
-   Kaplan-Meier Survival Curve

    -   Demonstrates the time where the event being modeled is expected
        to occur [@klein2005]

-   Forest Plots

    -   Visualizes the effects of each covariate in the model on the
        hazard ratio
:::

@@

## Checking Assumptions

## Model Results

@@

>>>>

@@

## Checking Assumptions {.scrollable}

::: panel-tabset
#### Independence

-   This assumption is met as the survival times for each individual
    engine in the data is independent of each other

#### Non-informative Censoring

-   This assumption is met as knowing there is censored data does not
    provide any new information about the survival probability

#### Linearity Assumption

-   This assumption is tested by looking at the Martingale residuals
    (Martingale Residuals = Observed Events - Expected Events) and
    determining if there is a slope of zero

-   Based on the below plots, this assumption is met.

```{r}
martingale <- resid(cox1.1, type = "martingale")

par(mfrow = c(4,4),
      mar = c(2,2,4,1))

plot(data_grouped$os1, martingale, main = "Martingale Residuals: os1")
abline(h = 0, col = "red")

plot(data_grouped$os2, martingale, main = "Martingale Residuals: os2")
abline(h = 0, col = "red")

plot(data_grouped$os3, martingale, main = "Martingale Residuals: os3")
abline(h = 0, col = "red")

plot(data_grouped$sm1, martingale, main = "Martingale Residuals: sm1")
abline(h = 0, col = "red")

plot(data_grouped$sm2, martingale, main = "Martingale Residuals: sm2")
abline(h = 0, col = "red")

plot(data_grouped$sm3, martingale, main = "Martingale Residuals: sm3")
abline(h = 0, col = "red")

plot(data_grouped$sm4, martingale, main = "Martingale Residuals: sm4")
abline(h = 0, col = "red")

plot(data_grouped$sm5, martingale, main = "Martingale Residuals: sm5")
abline(h = 0, col = "red")

plot(data_grouped$sm6, martingale, main = "Martingale Residuals: sm6")
abline(h = 0, col = "red")

plot(data_grouped$sm7, martingale, main = "Martingale Residuals: sm7")
abline(h = 0, col = "red")

plot(data_grouped$sm8, martingale, main = "Martingale Residuals: sm8")
abline(h = 0, col = "red")

plot(data_grouped$sm9, martingale, main = "Martingale Residuals: sm9")
abline(h = 0, col = "red")

plot(data_grouped$sm10, martingale, main = "Martingale Residuals: sm10")
abline(h = 0, col = "red")

plot(data_grouped$sm11, martingale, main = "Martingale Residuals: sm11")
abline(h = 0, col = "red")

plot(data_grouped$sm12, martingale, main = "Martingale Residuals: sm12")
abline(h = 0, col = "red")

plot(data_grouped$sm13, martingale, main = "Martingale Residuals: sm13")
abline(h = 0, col = "red")

plot(data_grouped$sm14, martingale, main = "Martingale Residuals: sm14")
abline(h = 0, col = "red")

plot(data_grouped$sm15, martingale, main = "Martingale Residuals: sm15")
abline(h = 0, col = "red")

plot(data_grouped$sm17, martingale, main = "Martingale Residuals: sm17")
abline(h = 0, col = "red")

plot(data_grouped$sm18, martingale, main = "Martingale Residuals: sm18")
abline(h = 0, col = "red")

plot(data_grouped$sm20, martingale, main = "Martingale Residuals: sm20")
abline(h = 0, col = "red")

plot(data_grouped$sm21, martingale, main = "Martingale Residuals: sm21")
abline(h = 0, col = "red")
```

#### Proportional Hazards

-   This assumption is tested by looking at the scaled Schoenfeld
    partial residuals (the value of the covariate - the expected value
    of the covariate at the time of failure) and examining the p-value

-   Based on the below table and plots, this assumption is met

```{r}
cph_ph <- cox.zph(cox1.1) %>%
  pluck("table") %>%
  as.data.frame() %>%
  tibble::rownames_to_column("variable") %>%
  as_tibble()
temp <- cph_ph %>% filter(variable %in% c("os1", "os2", "sm1", "sm2", "GLOBAL"))
kable(temp, format = "markdown")

create_ph_plots <- function(coxmodel, row, col){
  # Function takes argument that is a cox model from coxph()

  phresults <- cox.zph(coxmodel)

  # Define dimensions of visual
  par(mfrow = c(row, col),
      mar = c(2,2,4,1))

  # Loop through each covariate creating a plot and adding it to the phplot list
  for (i in 1:(length(phresults$table[,1])-1)){
    phplots <- list()
    phplots[i] <- plot(phresults[i], # Create plot for each iteration
                       main = rownames(phresults$table)[i]) # Title of each plot is the rowname from coxmodel$table

  }
  mtext("Schoenfeld Residuals vs. Time for Covariates", side = 3, line = -1.5, outer = TRUE, cex = 1) # Title for Visual
  par(mfrow = c(1,1)) # Reset plotting area
}

create_ph_plots(cox1.1, 3, 5)

```
:::

## Model Results {.scrollable}

::: panel-tabset
#### Evaluation

-   The CPH model was determined to be a valid model and has a
    concordance index of `{r} ci`

#### Calculations

-   Using the model, we can generate the coefficients and calculate the
    hazard rate, the cumulative hazard value, and the survival
    probability

```{r}
#| fig-align: center
data_grouped_reduced <- data_grouped %>% select(id, time, os1, os2, os3, sm1, sm2, sm3, sm4, sm5, sm6, sm7, sm8, sm9, sm10, sm11, sm12, sm13, sm14, sm15, sm17, sm18, sm20, sm21, status)

data_grouped_reduced$HazardRate <- predict(cox1.1,
        newdata = data_grouped_reduced,
        type = "risk")

data_grouped_reduced$SurvivalProbability <- predict(cox1.1, type = "survival", newdata = data_grouped_reduced)

print_data <- data_grouped_reduced %>% select(id, time, status, HazardRate, SurvivalProbability)

kable(head(print_data, n = 5), format = "markdown")

sf <- survfit(cox1.1)

ggsurvplot(fit = sf,
           data = data_grouped_reduced, risk.table = TRUE, fun = "cumhaz")$plot +
  ggtitle('Cumulative Hazard Over Time')

survtimes <- data.frame(Percent = c("100% Survival", "75% Survival,", "50% Survival", "25% Survival", "10% Survival"),
                        Time = c(sf$time[sf$lower < 1][1], 
                                    sf$time[sf$lower < .75][1], 
                                    sf$time[sf$lower < .5][1], 
                                    sf$time[sf$lower < .25][1], 
                                    sf$time[sf$lower < .1][1]))

ggsurvplot(fit = sf, data = data_grouped_reduced)$plot +
  ggtitle('Survival Probability Over Time')

```
:::

@@

-   Based on the model, NASA should not use their engines after 128
    iterations have passed

```{r}
kable(survtimes, format = "markdown")
```
